Relationships in the extremes and their influence on competition and coexistence

### 2021-09-24 paper/Paper.Rnw
some more edits to intro

However, if environmental responses become asymmetrically associated by, say, extreme rainfall events, coexistence can be altered. If the community is disturbed by frequent and long floods with some years of normalcy, the grass-forb/desert annual 
environmental response relationship may become more correlated in the higher values, a right-tailed ATA (figure \ref{fig:noise}C). Conversely, if the community is disturbed by frequent and long drought with some years of normalcy, the grass-forb/desert annual environmental response relationship may become more correlated in the lower values, a left-tailed ATA (figure \ref{fig:noise}A). 
The subsequent potential changes in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA). 
%Again, we see the possible consequences of ATA in Figure \ref{fig:pedagog}, i.e., less or no coexistence.

%***DAN: Leave the following for the next para
Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

@@

Impacts on competition and coexistence of climate change have been more commonly researched via studying changes in the mean of environmental variables; impacts of increased variance is less studied, although the rate of study has increased. \citep{Jentsch_2007}  %importance has been realized and variability is becoming more commonly studied now

@@

Although, the body of literature on the impact of increased environmental variability has expanded, event-based studies accounted for only one-fifth of experimental climate change research as of 2006 \citep{Jentch_2007}, the most commonly manipulated extremes being temperature and rainfall. Extreme drought and heat waves have shown impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. \cite{Fay_2003} experimentally manipulated rainfall for more frequent and intense precipitation while keeping constant the total volume of rain over Kansas tallgrass prairie and found that while some species had reduced fitness, overall community structure remained the same. In regard to invasive species, increased climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. 

@@

Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes, environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.

@@

%Move up the lottery model para from Theory to here, suture up the rift that this will produce in theory
To further demonstrate plausibility of ATA altering coexistence between species, we carried out simulations using a simple two-species model, the lottery model. 
Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, 
hence we move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. 
We introduce the lottery model and describe simulations based on it that clearly show that ATAs can influence 
species coexistence, though the details of this influence are the subject of the rest of the paper.
The lottery model is as follows.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year.
For each $i$, we assume the random variables $B_i(t)$ are independent and identically distributed (iid) through
time. We let $B_i = \exp(b_i)$, where $\left(b_1, b_2\right)$ is some bivariate random variable with the property 
that $b_i$ is normally distributed with mean $\mu_i$ and variance $\sigma^2$. 
However, we do \emph{not} assume that $\left(b_1, b_2\right)$ is a 
bivariate normal distribution. We will consider various distributions of $\left(b_1, b_2\right)$ with the above 
properties, corresponding to various tail association cases. Note that $N$ is constant through time.
%very succinct description of sims you do (mostly it's in the figure caption)
%1-2 sentences on the differences observed
We simulated two different communities up to \Sexpr{M} time steps to hopefully make it clear to the reader that ATA makes a difference for competition and coexistence between species (see Figure \ref{fig:pedagog}). The only distinction between the two simulations is that one depends on asymmetrically tail associated temporal dynamics and one does not. Markedly, the simulation with ATA has longer periods of one-species dominance than the simulation without, implying less recoverability when rare and thus less coexistence \citep{Chesson_2000} when environmentally dependent variables are asymmetrically tail associated.

>>>>

However, if environmental responses become asymmetrically associated by, say, extreme rainfall events, coexistence can be altered. If the community is disturbed by frequent and long floods with some years of normalcy, the grass-forb/desert annual 
environmental response relationship may become more correlated in the higher values, a right-tailed ATA (figure \ref{fig:noise}C). Conversely, if the community is disturbed by frequent and long drought with some years of normalcy, the grass-forb/desert annual environmental response relationship may become more correlated in the lower values, a left-tailed ATA (figure \ref{fig:noise}A). 
The subsequent potential changes in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA).

@@

Impacts on competition and coexistence of climate change have been more commonly researched via studying changes in the mean of environmental variables. 

@@

Although, the body of literature on the impact of increased environmental variability has expanded, event-based studies accounted for only one-fifth of experimental climate change research as of 2006 \citep{Jentsch_2007}, the most commonly manipulated extremes being temperature and rainfall. Extreme drought and heat waves have shown to impact competitive output, rearranging species composition and dominance of grass-forb communities \citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. \cite{Fay_2003} experimentally manipulated rainfall for more frequent and intense precipitation while keeping constant the total volume of rain over Kansas tallgrass prairie and found that while some species had reduced fitness, overall community structure remained the same. For invasive species, increased climate and resource variability allowed for greater success of invasibility in terms of density and competitive equilibrium \citep{Parepa_2013}. 

@@

Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes, environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, by using MCT, increased variability has been shown to promote coexistence in plants communities, via the storage effect \citep{Adler_2006,  Angert_2009}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence. 
Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

@@

%Move up the lottery model para from Theory to here, suture up the rift that this will produce in theory
To further demonstrate plausibility of ATA altering coexistence between species, we carried out simulations using a simple two-species model, the lottery model. 
Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, 
hence we move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. 
We introduce the lottery model and describe simulations based on it that clearly show that ATAs can influence 
species coexistence, though the details of this influence are the subject of the rest of the paper.
The lottery model is as follows.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year.
For each $i$, we assume the random variables $B_i(t)$ are independent and identically distributed (iid) through
time. We let $B_i = \exp(b_i)$, where $\left(b_1, b_2\right)$ is some bivariate random variable with the property 
that $b_i$ is normally distributed with mean $\mu_i$ and variance $\sigma^2$. 
However, we do \emph{not} assume that $\left(b_1, b_2\right)$ is a 
bivariate normal distribution. We will consider various distributions of $\left(b_1, b_2\right)$ with the above 
properties, corresponding to various tail association cases. Note that $N$ is constant through time.
%very succinct description of sims you do (mostly it's in the figure caption)
%1-2 sentences on the differences observed
We simulated two different communities up to \Sexpr{M} time steps to hopefully make it clear to the that ATA makes a difference for competition and coexistence between species (see Figure \ref{fig:pedagog} for more details). The only distinction between the two simulations is that one depends on asymmetrically tail associated temporal dynamics and one does not. Markedly, the simulation with ATA has longer periods of one-species dominance than the simulation without, implying less recoverability when rare and thus less coexistence \citep{Chesson_2000} when environmentally dependent variables are asymmetrically tail associated.
